---
title: "Final Project - Kernel and KNN"
author: "Alexander Gates"
date: "2026-04-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loading in the Data

```{r}
# Loading in the mlb data

swings <- read.csv("mlb_swings_2026.csv")
View(swings)
```

# Filtering using the filter function 
```{r}
library(tidyverse)

swings = swings |>
  mutate(hard_hit = if_else(launch_speed >= 95, 1, 0)) |>
  # remove deprecated / NA columns
  select(where(~!all(is.na(.)))) |>
  # remove irrelevant columns
  select(-c(game_date,
            player_name,
            batter,
            pitcher,
            des,
            game_type,
            home_team,
            away_team,
            hit_location,
            bb_type,
            game_year,
            on_3b,
            on_2b,
            on_1b,
            outs_when_up,
            inning,
            inning_topbot,
            hc_x,
            hc_y,
            hit_distance_sc,
            launch_angle,
            game_pk,
            fielder_2:fielder_9,
            estimated_ba_using_speedangle:launch_speed_angle,
            home_score:post_fld_score,
            delta_home_win_exp,
            delta_run_exp,
            estimated_slg_using_speedangle:age_bat_legacy,
            pitcher_days_since_prev_game,
            batter_days_since_prev_game,
            pitcher_days_until_next_game))

write_csv(swings, "mlb_swings_2026_reduced.csv")
```
The goal of this project is to classify whether a ball is a hard hit or not. A hard hit is defined as a ball with a launch speed of 95 mph or higher. We will use the `hard_hit` column as our target variable for classification. There will be two seperate analysis conducted, one of them will be using a kernel method and the other will be using KNN. For both of these methods, we will need to filter the data to just have the relevant variables for our analysis, as we will focus on just pitching related variables and the target variable (hard_hit).

First we will filter the data to just have the relevant variables for our analysis, as we will focus on just pitching related variables and the target variable (hard_hit). We will use the `select` function from the `dplyr` package to select only the relevant columns from the `swings` dataset.

```{r}
# Filtering to just have pitching related variables and the target variable (hard_hit)
swings_pitch = swings |>
  select(pitch_type, release_speed, effective_speed,
release_pos_x, release_pos_y, release_pos_z,
release_extension, arm_angle,
release_spin_rate, spin_axis,
pfx_x, pfx_z,
vx0, vy0, vz0, ax, ay, az,
plate_x, plate_z, zone,
balls, strikes, p_throws,
sz_top, sz_bot,
n_thruorder_pitcher, pitch_number, hard_hit)
```

Now, we will need to convert the `pitch_type` variable into a factor variable, as it is currently a character variable. We will use the `as.factor` function to convert it into a factor variable, as well as `p_throws` into a factor variable as well, since it is also a character variable. This will allow us to use these variables in our classification models.

We will also remove `hard_hit` NA's from the dataset, as they can cause issues with our classification models. We will use the `na.omit` function to remove any rows with NA's from the dataset.

```{r}
# filtering out NA's and converting character variables to factors
swings_pitch <- swings_pitch |>
  mutate(
    pitch_type = as.factor(pitch_type),
    p_throws = as.factor(p_throws),
    hard_hit = factor(hard_hit, levels = c(0, 1))
  ) |>
  filter(!is.na(hard_hit))|>
  na.omit()




```


```{r}
set.seed(123) # Setting a seed for reproducibility

# Splitting the data into a training set and a test set
train_index = sample(1:nrow(swings_pitch), 0.8 * nrow(swings_pitch))
train_data = swings_pitch[train_index, ]
test_data = swings_pitch[-train_index, ]
```


```{r}
library(e1071)

# Fitting an SVM model to the training data
svm_model = svm(hard_hit ~ ., data = train_data, kernel = "radial")
# Predicting on the test data
svm_predictions = predict(svm_model, test_data)

# Evaluating the performance of the SVM model
svm_confusion_matrix = table(test_data$hard_hit, svm_predictions)
svm_accuracy = sum(diag(svm_confusion_matrix)) / sum(svm_confusion_matrix)
svm_accuracy

table(svm_predictions, test_data$hard_hit)
```


```{r}
library(caret)
library(e1071)

weight_values <- c(1, 2, 3, 4, 5)

results <- data.frame(weight = numeric(), accuracy = numeric())

for (w in weight_values) {
  
  weights <- ifelse(train_data$hard_hit == "1", w, 1)
  
  svm_model <- train(
    hard_hit ~ .,
    data = train_data,
    method = "svmRadial",
    preProcess = c("center", "scale"),
    tuneGrid = grid,
    trControl = trainControl(method = "cv", number = 3),
    weights = weights
  )
  
  pred <- predict(svm_model, test_data)
  
  acc <- sum(pred == test_data$hard_hit) / length(pred)
  
  results <- rbind(results, data.frame(weight = w, accuracy = acc))
}

results



```
```{r}
ggplot(results, aes(x = weight, y = accuracy)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text(aes(label = round(accuracy, 3)), vjust = -0.8) +
  ylim(0, 1) +
  theme_minimal() +
  labs(
    title = "SVM Accuracy vs Class Weight (Class 1)",
    x = "Weight for Minority Class",
    y = "Accuracy"
  )
```

# KNN Baseline
```{r}
library(caret)
set.seed(123)

knn_baseline <- train(
  hard_hit ~ .,
  data = train_data,
  method = "knn",
  preProcess = c("center", "scale"),
  tuneGrid = expand.grid(k = 5),   # pick a reasonable default
  trControl = trainControl(method = "none")
)

knn_baseline


pred_base <- predict(knn_baseline, test_data)

conf_mat_base <- table(test_data$hard_hit, pred_base)
conf_mat_base

accuracy_base <- sum(diag(conf_mat_base)) / sum(conf_mat_base)
accuracy_base
```


# KNN Selection
```{r}

set.seed(123)

knn_grid <- expand.grid(
  k = c(3, 5, 7, 9, 11, 13, 15, 30, 50, 100)  
)

knn_model <- train(
  hard_hit ~ .,
  data = train_data,
  method = "knn",
  preProcess = c("center", "scale"),  # VERY important for KNN
  tuneGrid = knn_grid,
  trControl = trainControl(
    method = "cv",
    number = 5
  )
)

knn_model
plot(knn_model)

pred_knn <- predict(knn_model, test_data)

conf_mat_knn <- table(test_data$hard_hit, pred_knn)
conf_mat_knn

accuracy_knn <- sum(diag(conf_mat_knn)) / sum(conf_mat_knn)
accuracy_knn
```

```{r}
# comparison of all 4 plots

# svm_model

svm_model
# choose the best for svm_2 

# best choice is weight = 3
weights <- ifelse(train_data$hard_hit == "1", 3, 1)
  
svm_model_weighted <- train(
    hard_hit ~ .,
    data = train_data,
    method = "svmRadial",
    preProcess = c("center", "scale"),
    tuneGrid = grid,
    trControl = trainControl(method = "cv", number = 3),
    weights = weights)

# make the accuracy for svm_2
pred_svm_weighted <- predict(svm_model_weighted, test_data)
conf_mat_svm_weighted <- table(test_data$hard_hit, pred_svm_weighted)
accuracy_svm_weighted <- sum(diag(conf_mat_svm_weighted)) / sum(conf_mat_svm_weighted)






# knn_baseline

knn_baseline

# neighbors = 3
knn_best <- train(
  hard_hit ~ .,
  data = train_data,
  method = "knn",
  preProcess = c("center", "scale"),
  tuneGrid = expand.grid(k = 3),   # pick a reasonable default
  trControl = trainControl(method = "none")
)

# accuracy for knn_best

pred_knn_best <- predict(knn_best, test_data)
conf_mat_knn_best <- table(test_data$hard_hit, pred_knn_best)
accuracy_knn_best <- sum(diag(conf_mat_knn_best)) / sum(conf_mat_knn_best)


# data frame of all 4 accuracies
results <- data.frame(
  Model = c("SVM Baseline", "SVM Weighted", "KNN Baseline", "KNN Best"),
  Accuracy = c(svm_accuracy, accuracy_svm_weighted, accuracy_base, accuracy_knn_best)
)

# plot
library(ggplot2)

ggplot(results, aes(x = Model, y = Accuracy)) +
  geom_bar(stat = "identity") +
  ylim(0, 1) +
  labs(title = "Model Accuracy Comparison",
       y = "Accuracy",
       x = "Model") +
  theme_minimal()




```
```{r}
# Table of all 4 conf_mat

conf_matrices <- list(
  "SVM Baseline"  = svm_confusion_matrix,
  "SVM Weighted"  = conf_mat_svm_weighted,
  "KNN Baseline"  = conf_mat_base,
  "KNN Best"      = conf_mat_knn_best
)

conf_matrices

library(dplyr)

cm_to_df <- function(cm, model_name) {
  df <- as.data.frame(cm)
  colnames(df) <- c("Actual", "Predicted", "Freq")
  df$Model <- model_name
  return(df)
}

all_cm <- bind_rows(
  cm_to_df(svm_confusion_matrix, "SVM Baseline"),
  cm_to_df(conf_mat_svm_weighted, "SVM Weighted"),
  cm_to_df(conf_mat_base, "KNN Baseline"),
  cm_to_df(conf_mat_knn_best, "KNN Best")
)

all_cm

library(ggplot2)


ggplot(all_cm, aes(x = Predicted, y = Actual, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white") +
  facet_wrap(~ Model) +
  theme_minimal() +
  labs(title = "Confusion Matrices Comparison",
       x = "Predicted",
       y = "Actual")
```

